An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory

### 2016-07-09 Papers/2016/KI/main.tex
Improve text about the inconsistency

In the typical workflow during an attempt to prove a conjecture with a
theorem prover, it is customary to check the consistency of the axioms
first. For if the axioms are inconsistent, anything (including the
conjecture) would be trivially derivable in classical logic (\emph{ex
  falso quodlibet}). Surprisingly, when this routine check was
performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover
claimed that the axioms were inconsistent. Unfortunately, the
refutation generated by \textsc{Leo-II} was barely human-readable. The
text file was 153 lines\footnote{Long lines with an average of 184
  characters per line.} long and used machine-oriented calculus
(higher-order resolution \cite{W47}) and syntax (TPTP THF
\cite{J22}).

% Part of the file is displayed in Fig.~\ref{LEO-Proof}.

% \begin{figure*}
% \centerline{\includegraphics[width=\textwidth]{./LEO-Proof.png}}
% \caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
%   substitutions (e.g. with the empty property) are highlighted. In the
% red part (see $\boldsymbol{\leftarrow}$), property variable $\texttt{SV8}$ has been instantiated with the
% $\lambda \texttt{SV16}_{\mu}. \lambda \texttt{SV17}_{\iota}. \bot$, i.e., the (lifted) empty property.
% } \label{LEO-Proof}
% \end{figure*}

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets).
%
Note that the terms for the empty property\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.} ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with full functional and Boolean extensionality as given
here. Nevertheless, some philosophers\footnote{Private communication with Andr\'e Fuhrmann.} may actually prefer the use of
self-difference over the empty property in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference.

>>>>

In the typical workflow during an attempt to prove a conjecture with a
theorem prover, it is customary to check the consistency of the axioms
first. For if the axioms are inconsistent, anything (including the
conjecture) would be trivially derivable in classical logic (\emph{ex
  falso quodlibet}). Surprisingly, when this routine check was
performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover
claimed that the axioms were inconsistent. Unfortunately, the
refutation generated by \textsc{Leo-II} was barely human-readable. The refutation was based on machine-oriented inference rules (the higher-order resolution calculus \cite{W47}), and the
text file had 153 lines (with an average of 184
  characters per line) and used a machine-oriented syntax (TPTP THF
\cite{J22}).

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets).
%
Note that the terms for the empty property\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.} ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in a logic setting
with functional and Boolean extensionality as assumed
here. Nevertheless, some philosophers\footnote{Private communication with Andr\'e Fuhrmann.} may actually prefer the use of
self-difference over the empty property in
the analysis below. However, for the proof to go through it is
irrelevant which notion we use and the reader may simply replace the
empty property by self-difference.
